---
title: "Visualizing government research and development spending"
format: 
  html:
    toc: true
    toc-location: right
    number-depth: 4
    theme: cerulean
    df-print: kable
execute:
  echo: true
editor: source
number-sections: true
embed-resources: true
---

## Background information

The version of the data that we are using comes from "Tidy Tuesday" ([link](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-02-12#federal-research-and-development-spending-by-agency)), but the original data source is from the American Association for the Advancement of Science ([link](https://www.aaas.org/programs/r-d-budget-and-policy/historical-trends-federal-rd)). The version of the data we are using has already been cleaned by Tom Mock (a Tidy Tuesday facilitator), and he shares the code he used to clean the data [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-02-12#federal-research-and-development-spending-by-agency).

## Load in the data

```{r, message=FALSE}
library(tidyverse)
```

First, we load in the three files.

```{r, message=FALSE}
fed_spending <- read_csv("data/fed_rd.csv")
energy_spending <- read_csv("data/energy_spend.csv")
climate_spending <- read_csv("data/climate_spend.csv") 
```

Let's take a look at the first few rows of each dataset.

```{r}
head(fed_spending)
```

```{r}
head(energy_spending)
```

```{r}
head(climate_spending)
```

## Prepare the data

Next, we want to prepare the data (this is essentially pre-processing the data for our visualization analysis below). To simplify the data, we make the judgment call to aggregate/summarize the spending across all departments for each year.

```{r}
# aggregate the energy spending by year
energy_by_year <- energy_spending %>%
  group_by(year) %>%
  summarise(energy_spending = sum(energy_spending))
```

```{r}
# aggregate the climate spending by year
climate_by_year <- climate_spending %>%
  group_by(year) %>%
  summarise(climate_spending = sum(gcc_spending))
```

```{r}
# aggregate the federal spending by year
fed_by_year <- fed_spending %>%
  group_by(year) %>%
  summarise(total_rd_budget = sum(rd_budget),
            total_spending = unique(total_outlays),
            discretionary_spending = unique(discretionary_outlays),
            gdp = unique(gdp)) 
```

And then we can combine all three datasets together using `left_join()`, filter to the year 2000 onwards, and scale each of the relevant variables by one million.

```{r}
govt_spending <- fed_by_year %>%
  # join the energy and climate spending to the federal spending data
  left_join(energy_by_year, by = "year") %>%
  left_join(climate_by_year, by = "year") %>%
  # filter to the year 2000 onwards
  filter(year >= 2000) %>%
  # divide all relevant variables by 1000000
  mutate_at(vars(gdp,
                 total_rd_budget,
                 total_spending,
                 discretionary_spending,
                 energy_spending,
                 climate_spending),
            function(x) x / 1000000)
```

## Exercise:

Now it's your turn to re-create the figures specified in the exercise.

You will probably want to write some code that starts like this:

```{r}
govt_spending %>%
  ggplot() 
  
```
